---
title: "Herbivory coding"
format: html
editor: visual
execute: 
  cache: true
  message: true
  warning: true
  echo: false
toc: true
toc-location: left
code-fold: true
code-tools: true
---

## Import and wrangle data

```{r}
#| label: import data 
## load libraries

library(pacman)
pacman::p_load(tidyverse, readxl, here, skimr, overviewR, ggmap)       ## readxl is needed to load excel files into R
                                              ## the `here` package helps with file paths


path <- here::here("data")

xls <- list.files(path, "xls", full.names = TRUE)

data <- readxl::read_xlsx(xls[2])

data <- data |>
  janitor::clean_names()

```

Lets look at the data

```{r}
#| label: descriptives
nrow <- nrow(data)
ncol <- ncol(data)

skimr::skim(data)
```

-   The data set has `r nrow` rows and `r ncol` columns

```{r}

data |>
  count(europe)
```

## Mapping

To map locations of studies we need to

1.  Extract the lat longs
2.  Convert them to decimal coordinates

Lets look at the different ways in which lat-longs are recorded in the data so we can create string patterns to extract and convert the values.

The basic pattern is degrees and minutes, some have seconds - this can be represented as 1-2 digits followed by a variable number of non-digits repeated 2 or 3 times and with a compass point at the beginning or the end.

Looking at the unique contents of the `latitude_n_s` field we can also see it contains non lats (a DOI) and also other numbers which we need to remove or exclude.

```{r}
#| label: unique lat-longs
#| eval: false
data$latitude_n_s[1] |>
  str_extract(pattern2)
```

```{r}

#| label: create regex patterns 

#pattern1 <- "\\d{1,2}\\.\\d{1,2}"                                  ## e.g. N54:41:18 
pattern2 <- "\\d{1,2}\\D*\\d{1,2}\\D*\\d{1,2}\\D*[NSEW]"             ## e.g. 52 8' 12\"N
pattern3 <- "\\d{1,2}\\D*\\d{1,2}\\D*[NS]"                         ## e.g. 52°86' N
pattern2a <- "\\d{1,2}\\D*\\d{1,2}\\.\\d{1,2}\\.\\d{1,2}\\D*[NSEW]" 
pattern2b <- "\\d{1,2}\\D*\\d{1,2}\\.\\d{1,2}\\D*[NSEW]"## eg 52 25' 43.76"N
pattern4 <- "[EW]\\d{1,2}\\D*\\d{1,2}\\D*\\d{1,2}"
pattern5 <- "\\d{1,2}\\D*\\d{1,2}\\D*\\d{1,2}\\D*[EW]" 
pattern6 <- "\\d{1,2}\\D*\\d{1,2}\\D*[EW]"
pattern7 <- "[NS]\\d{1,2}[[:punct:]]\\d{1,2}[[:punct:]]\\d{1,2}"
pattern8 <- "[EW]\\d{1,2}[[:punct:]]\\d{1,2}[[:punct:]]\\d{1,2}"


```

We can use these patterns to isolate the lat-longs, extract the numerical data and calculate the decimal coordinates.

```{r}
#| label: dry run
library(stringi)
lat_long <- data|>
  select(item, latitude_n_s, longitude_e_w) |>
  mutate_all(~str_remove_all(.,"20 sites")) |>
  mutate_all(~str_remove_all(., "Hill.\\d")) |>
  mutate_all(~str_remove_all(., "NY\\D*\\d{6}")) |>
  mutate_all(~str_remove_all(., "(state in info)\r\n") )|>
  mutate(latitude_n_s = str_remove(latitude_n_s, pattern8), 
         longitude_e_w = str_remove(longitude_e_w, pattern7))

lat_long


ll <- lat_long |>
  mutate(lat = str_extract(latitude_n_s, pattern2a), 
         lat = ifelse(is.na(lat), str_extract(latitude_n_s, pattern2), lat), 
         lat = ifelse(is.na(lat), str_extract(latitude_n_s, pattern7), lat), 
         long = str_extract(longitude_e_w, pattern2),
         long = ifelse(is.na(long), str_extract(longitude_e_w, pattern2b), long),
         long = ifelse(is.na(long), str_extract(longitude_e_w, pattern8), long), 
         long = ifelse(is.na(long), str_extract(longitude_e_w, pattern6), long))

ll |>
  filter(str_detect(item, "Medi"))

```

```{r}
#| label: calculate decimal lat-longs
#| message: false
#| 
#lats <- 
  
lat_long |>
  filter(str_detect(item, "^D"))
```

```{r}
  
trial <- ll |> 
  mutate(lat1 = str_replace_all(lat, "\\D", " "),
         long1 = str_replace_all(long, "\\D", " ")) |>
  mutate(latlen = str_length(lat1)) 

points <- trial |> 
  select(item, lat, long) |>
  mutate(compass_lat = str_extract(lat, "[NS]"), 
         compass_long = str_extract(long, "[EW]"), 
         lat1 = str_extract_all(lat, "\\d{1,2}"), 
         long1 = str_extract_all(long, "\\d{1,2}")) |>
  drop_na() |>
  unnest("lat1") |>
  group_by(item) |>
  mutate(lat_id = row_number()) |>
  filter(!lat_id %in% c(3,4)) |>
  mutate(lat1 = as.numeric(lat1), 
         dec_lat = ifelse(str_detect(lat, "\\d{1,2}\\.\\d{1,2}"), lat1[1] + lat1[2]/100, lat1[1] + lat1[2]/60)) |>
  select(-c(lat1, lat_id)) |>
  unnest("long1") |>
  mutate(long_id = row_number()) |>
  filter(long_id %in% c(1,2)) |> 
  mutate(long1 = as.numeric(long1),
         dec_long = ifelse(str_detect(long, "\\d{1,2}\\.\\d{1,2}"), long1[1] + long1[2]/100, long1[1] + long1[2]/60), 
         dec_long = ifelse(compass_long == "W", dec_long * -1, dec_long), 
         dec_lat = ifelse(compass_long == "S", dec_lat * -1, dec_lat)) |>
  select(item, dec_lat, dec_long) |>
  distinct()
  
  

```

```{r}

## temp fix of daaz while working out a better solution

daaz_lat <- 51 + 41/60
daaz_long <- 2 + 37/60

points[23, "dec_lat"] <- daaz_lat
points[23, "dec_long"] <- daaz_long

points
```

```{r}
library(sf)
  coords <- points |>
  st_as_sf(coords = c(x = "dec_long", y = "dec_lat"), crs = 4326)
  
coords
```

## Interactive map

```{r}
library(mapview);library(sp);library(ggspatial)

mapview(coords, legend = FALSE)
```

## Static map

```{r}

coords |>
  ggplot() +
  annotation_map_tile("cartolight") +
  geom_sf() +
  theme_void()
  

```
